---
title: "Inferential Analysis on Tooth Growth by Supplement Types and Dosages"
author: "Jiachang (Ernest) Xu"
date: "6/12/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project is a breif graphical and numericla exploration of the dataset, by performing the techniques of confidence interval and hypothesis testing.

## Data Processing and Cleaning

```{r summary, echo=TRUE}
library(datasets)
require(ggplot2)
require(RColorBrewer)
require(grDevices)

data(ToothGrowth)
attach(ToothGrowth)

# first look: 3 variables and 60 observations
str(ToothGrowth)
summary(ToothGrowth)

# convert $dose to factors
unique(ToothGrowth$dose)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

```

## Exploratory Analysis

This section examines the relationship between dose size, dose type, and tooth length

```{r, echo=TRUE, fig.height=3}
require(ggplot2)
require(gridExtra)

theme <- theme(
    panel.background = element_rect(fill = "lightgrey", colour = "lightgrey", size = 0.5, linetype = "solid"),
    panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour = "white"), 
    panel.grid.minor = element_line(size = 0.25, linetype = 'solid', colour = "white")
)

plot1 <- ggplot(aes(x = factor(dose), y = len), data = ToothGrowth) +
                geom_boxplot(aes(fill = factor(dose))) + 
                theme + scale_fill_brewer(palette="GnBu") + 
                labs(title = "Figure 1")

plot2 <- ggplot(aes(x = supp, y = len), data = ToothGrowth) + 
                geom_boxplot(aes(fill = supp)) + 
                theme + scale_fill_brewer(palette="PuOr") + 
                labs(title = "Figure 2")

grid.arrange(plot1, plot2, ncol=2)
```

**Figure 1** shows that tooth length increases as the doseage increases.

**Figure 2** shows that Orange Juice is more effective than Vitamin C, including all doseage levels.

```{r, echo=TRUE, fig.height=3}
require(ggplot2)
require(gridExtra)

plot3 <- ggplot(aes(x = supp, y = len), data = ToothGrowth) + 
                geom_boxplot(aes(fill = supp)) + 
                facet_wrap(~ dose) + theme + 
                scale_fill_brewer(palette="YlOrRd") + 
                labs(title = "Figure 3: Tooth Growth due to Two Supplements by Incremental Dosages") + 
                annotate("text", x = 1.5, y = 2, label = c("p = .03","p = .0005","p = .5"))
plot3
```

**Figure 3** shows that Orange Juice is more effective than Vitamin C only in the first two doseage, but they are equally effective in the third doseage.

## Hypothesis Testing

**First**, we need to reformat the data by using the split() function. Therefore, we can do hypothesis testing more conviently from the reformatted dataset.

```{r split, echo=TRUE}
# split the data frame by dose and supplement type
split_tooth <- split(ToothGrowth, f = list(ToothGrowth$dose, ToothGrowth$supp))
```

**Second**, aggregations show the sample mean and standard deviation.

```{r aggregation, echo=TRUE}
aggregate(len, list(supp, dose), mean)
aggregate(len, list(supp, dose), sd)
```

**Third**, we perform hypothesis testing at the **5%** significance level. Each p-value correponds to the question immediately above it

#####Test 1: Is Orange Juice more effective than vitamin C across doses? <span style="color:green">YES</span>
```{r test 1, echo=TRUE}
t.test(c(split_tooth[[1]]$len, split_tooth[[2]]$len, split_tooth[[3]]$len), 
       c(split_tooth[[4]]$len, split_tooth[[5]]$len, split_tooth[[6]]$len), 
       alternative = "greater")$p.value
```

#####Test 2: Is Orange Juice more effective than Vitamin C for a .5 dose? <span style="color:green">YES</span>
```{r test 2, echo=TRUE}
t.test(split_tooth[[1]]$len, split_tooth[[4]]$len, 
       alternative = "greater")$p.value
```

#####Test 3: Is Orange Juice more effective than Vitamin C for a 1 dose? <span style="color:green">YES</span>
```{r test 3, echo=TRUE}
t.test(split_tooth[[2]]$len, split_tooth[[5]]$len, 
       alternative = "greater")$p.value
```

#####Test 4: Is Orange Juice more effective than Vitamin C for a 2 dose? <span style="color:red">INCONCLUSIVE</span>
```{r test 4, echo=TRUE}
t.test(split_tooth[[3]]$len, split_tooth[[6]]$len, 
       alternative = "greater")$p.value
```

#####Test 5: Is a 2 dose (any supplement type) more effective than a .5 dose? <span style="color:green">YES</span>
```{r test 5, echo=TRUE}
t.test(c(split_tooth[[1]]$len, split_tooth[[4]]$len), 
       c(split_tooth[[2]]$len, split_tooth[[5]]$len), 
       alternative = "less")$p.value
```

#####Test 6: Is a 2 dose (any supplement type) more effective than a .5 dose? <span style="color:green">YES</span>
```{r test 6, echo=TRUE}
t.test(c(split_tooth[[1]]$len, split_tooth[[4]]$len), 
       c(split_tooth[[3]]$len, split_tooth[[6]]$len), 
       alternative = "less")$p.value
```

#####Test 7: Is a 2 dose (any supplement type) more effective than a .5 dose? <span style="color:green">YES</span>
```{r test 7, echo=TRUE}
t.test(c(split_tooth[[2]]$len, split_tooth[[5]]$len), 
       c(split_tooth[[3]]$len, split_tooth[[6]]$len), 
       alternative = "less")$p.value
```
